---
title: "RandomForest"
output: html_document
date: "2025-10-11"
---

```{r setup, include=FALSE}
library(tidyverse)
library(hoopR)
library(ranger)
library(xgboost)
```


```{r data gathering, eval=FALSE}
library(hoopR)
library(parallel)
library(tidyverse)
check_is_team = function(p, team1, team2){
  return(as.character(p) == as.character(team1) | as.character(p) == as.character(team2))
}

SEASON_NAME = 2024
get_pbp_game_sequence_data = function(game_i){
  pbp = hoopR::nba_data_pbp(game_id=game_i) %>% select(period, event_num, clock, description,offense_team_id,locX,locY,player1_id,player2_id,player3_id,home_score,away_score)
  pbp$event_num = as.numeric(pbp$event_num)
  pbp = pbp %>% arrange(event_num)
  players = hoopR::nba_boxscorescoringv2(game_id=game_i)$sqlPlayersScoring %>% select(TEAM_ID, PLAYER_ID) %>% rename(team_id = TEAM_ID, player_id = PLAYER_ID) %>% arrange(team_id) %>% unique()
  teams = players %>% select(team_id) %>% unique() %>% pull('team_id')
  team = players[1,'team_id'] %>% pull()
  player = players[1,'player_id'] %>% pull()
  all_shots = hoopR::nba_shotchartdetail(team_id = team, season=year_to_season(SEASON_NAME), player_id=player, game_id=game_i)$Shot_Chart_Detail
  for(i in 2:nrow(players)){
    team = players[i,'team_id'] %>% pull()
    player = players[i,'player_id'] %>% pull()
    tryCatch({
      player_shots = hoopR::nba_shotchartdetail(team_id = team, season=year_to_season(SEASON_NAME), player_id=player, game_id=game_i)$Shot_Chart_Detail
      all_shots = all_shots %>% rbind(player_shots)
    }, error=function(e){})
    Sys.sleep(1.5)
  }
  all_shots$GAME_EVENT_ID = as.numeric(all_shots$GAME_EVENT_ID)
  all_shots = all_shots %>% arrange(GAME_EVENT_ID) %>% rename(event_num = GAME_EVENT_ID, shot_loc_x = LOC_X, shot_loc_y = LOC_Y) %>% select(event_num,ACTION_TYPE,SHOT_ZONE_AREA, SHOT_ZONE_RANGE, SHOT_DISTANCE,shot_loc_x,shot_loc_y,SHOT_MADE_FLAG)
  colnames(all_shots) = tolower(colnames(all_shots))
  print(colnames(all_shots))
  pbp_w_shots = pbp %>% left_join(all_shots) %>% rename(poss_loc_x = locX, poss_loc_y = locY) 
  ## start building sequence data
  ## sequences start: defensive rebound, offensive rebound, steal, timeout
  ## sequences end: shot (make, miss), turnover
  
  ## init_action: defensive rebound, offensive rebound, steal, timeout
  ## end_action: shot (make or miss), turnover
  ## init_player_id: player responsible for start of sequence (i.e. rebounder, stealer, etc.)
  ## end_player_id: player responsible for ending possess
  ## init_play_loc_x: starting x location of sequence
  ## init_play_loc_y: starting y location of sequence
  ## end_play_loc_x: ending x location of sequence
  ## end_play_loc_y: ending y location of sequence
  ## shot_distance?:
  ## shot_zone?: 
  ## success?: 1 if make, 0 if not
  pbp_w_shots = pbp_w_shots %>% filter(player1_id > 0)
  sequences = tibble() %>% mutate(
    sequence_id = c(1),
    offense_team_id = c(0),
    init_action = c(''),
    end_action = c(''),
    success = as.numeric(c(0)),
    init_player_id = c(''),
    end_player_id = c(''),
    init_play_loc_x = as.numeric(c(0)),
    init_play_loc_y = as.numeric(c(0)),
    end_play_loc_x = as.numeric(c(0)),
    end_play_loc_y = as.numeric(c(0)),
    shot_distance = as.numeric(c(0)),
    shot_zone = c(''),
    shot_result = as.numeric(0),
    shot_type = as.character(''),
    shot_was_tpa = as.numeric(0),
  )
  curr_init_action = ''
  curr_team = ''
  curr_init_player_id = ''
  curr_init_play_loc_x = 0
  curr_init_play_loc_y = 0
  
  for(i in 1:nrow(pbp_w_shots)){
    tryCatch({
      if(i > 1){
        if(curr_init_action == ''){
          if(tolower(pbp_w_shots[i, 'description']) %>% str_detect('rebound') | (sequences[nrow(sequences),] %>% pull(end_action) == 'shot' & sequences[nrow(sequences),] %>% pull(success) == 0)){
            curr_team = pbp_w_shots[i,] %>% pull('player1_id')
            if(!check_is_team(curr_team, teams[1], teams[2]))
            {
              curr_team = players %>% filter(player_id == curr_team) %>% pull('team_id')
            }
            curr_init_action = 'rebound'  
            curr_init_player_id = pbp_w_shots[i,] %>% dplyr::pull('player1_id')
          }
          else if(tolower(pbp_w_shots[i-1, 'description']) %>% str_detect('steal')){
            curr_init_action = 'steal'
            curr_init_player_id = pbp_w_shots[i-1,] %>% dplyr::pull('player3_id')
          }
          else if(tolower(pbp_w_shots[i-1, 'description']) %>% str_detect('turnover')){
            curr_init_action = 'violation'
            curr_init_player_id = pbp_w_shots[i,] %>% dplyr::pull('player1_id')
          }
          else if(tolower(pbp_w_shots[i-1, 'description']) %>% str_detect('timeout')){
            curr_init_action = 'timeout'
            curr_init_player_id = ''
          }
          else{
            curr_init_action = 'inbound'
            curr_init_player_id = pbp_w_shots[i,] %>% dplyr::pull('player1_id')
          }
          curr_init_play_loc_x = pbp_w_shots[i,] %>% dplyr::pull('poss_loc_x')
          curr_init_play_loc_y = pbp_w_shots[i,] %>% dplyr::pull('poss_loc_y')
        }
        else {
          end_success = 0
          end_loc_x = pbp_w_shots[i,] %>% pull('poss_loc_x')
          end_loc_y = pbp_w_shots[i,] %>% pull('poss_loc_y')
          end_shot_distance = -1
          end_shot_type = ''
          end_shot_zone = -1
          end_shot_result = -1
          end_shot_was_tpa = -1
          if(tolower(pbp_w_shots[i, 'description']) %>% str_detect('shot')){
            end_action = 'shot'
            end_success = pbp_w_shots[i,] %>% pull('shot_made_flag')
            end_loc_x = pbp_w_shots[i,] %>% pull('shot_loc_x')
            end_loc_y = pbp_w_shots[i,] %>% pull('shot_loc_y')
            end_shot_distance = pbp_w_shots[i,] %>% pull('shot_distance')
            end_shot_zone = pbp_w_shots[i,] %>% pull('shot_zone_area')
            if(tolower(pbp_w_shots[i,] %>% pull('description')) %>% str_detect('3pt')){
              end_shot_was_tpa = 1
              if(end_success == 1){
                end_shot_result = 3
              }
              else {
                end_shot_result = 0
              }
            }
            else{
              end_shot_was_tpa = 0
              if(end_success == 1){
                end_shot_result = 2
              }
              else {
                end_shot_result = 0
              }
            }
            
            end_shot_type = tolower(pbp_w_shots[i,] %>% pull('action_type'))
          }
          else if(tolower(pbp_w_shots[i, 'description']) %>% str_detect('turnover')){
            end_action = 'turnover'
          }
          end_player = pbp_w_shots[i,] %>% pull('player1_id')
          sequences = sequences %>% 
            add_row(sequence_id=nrow(sequences),
                    offense_team_id = as.double(curr_team),
                    init_action = curr_init_action,
                    end_action = end_action,
                    success=as.numeric(end_success),
                    init_player_id = as.character(curr_init_player_id),
                    end_player_id = as.character(end_player),
                    init_play_loc_x = as.numeric(curr_init_play_loc_x),
                    init_play_loc_y = as.numeric(curr_init_play_loc_y),
                    end_play_loc_x= as.numeric(end_loc_x),
                    end_play_loc_y = as.numeric(end_loc_y),
                    shot_distance = as.numeric(end_shot_distance),
                    shot_zone = as.character(end_shot_zone),
                    shot_result = as.numeric(end_shot_result),
                    shot_type = as.character(end_shot_type),
                    shot_was_tpa = as.numeric(end_shot_was_tpa))
          curr_init_action = '' 
        }
      }
      else{
        curr_init_action = 'tipoff'
        curr_init_play_loc_x = 0
        curr_init_play_loc_y = -80
        p = pbp_w_shots[i,] %>% pull(player1_id)
        curr_team = as.double(players[players$player_id == p,] %>% pull('team_id'))
        if(tolower(pbp_w_shots[i, 'description']) %>% str_detect('shot')){
          end_action = 'shot'
          end_success = pbp_w_shots[i,] %>% pull('shot_made_flag')
          end_loc_x = pbp_w_shots[i,] %>% pull('shot_loc_x')
          end_loc_y = pbp_w_shots[i,] %>% pull('shot_loc_y')
          end_shot_distance = pbp_w_shots[i,] %>% pull('shot_distance')
          end_shot_zone = pbp_w_shots[i,] %>% pull('shot_zone_area')
          end_shot_result = 0
          end_shot_type = ''
          end_shot_was_tpa = -1
          if(tolower(pbp_w_shots[i,] %>% pull('description')) %>% str_detect('3pt')){
            end_shot_was_tpa = 1
            if(end_success == 1){
              end_shot_result = 3
            }
            else {
              end_shot_result = 0
            }
          }
          else{
            end_shot_was_tpa = 0
            if(end_success == 1){
              end_shot_result = 2
            }
            else {
              end_shot_result = 0
            }
          }
          
          end_shot_type = tolower(pbp_w_shots[i,] %>% pull('action_type'))
          end_player = pbp_w_shots[i,] %>% pull('player1_id')
          sequences = sequences %>% 
            add_row(sequence_id=nrow(sequences),
                    offense_team_id = as.double(curr_team),
                    init_action = curr_init_action,
                    end_action = end_action,
                    success=as.numeric(end_success),
                    init_player_id = as.character(curr_init_player_id),
                    end_player_id = as.character(end_player),
                    init_play_loc_x = as.numeric(curr_init_play_loc_x),
                    init_play_loc_y = as.numeric(curr_init_play_loc_y),
                    end_play_loc_x= as.numeric(end_loc_x),
                    end_play_loc_y = as.numeric(end_loc_y),
                    shot_distance = as.numeric(end_shot_distance),
                    shot_zone = as.character(end_shot_zone),
                    shot_result = as.numeric(end_shot_result),
                    shot_type = as.character(end_shot_type),
                    shot_was_tpa = as.numeric(end_shot_was_tpa))
          curr_init_action = ''
        }
        else if(tolower(pbp_w_shots[i, 'description']) %>% str_detect('timeout')){
          end_action = 'timeout'
          end_success = 0
          end_loc_x = pbp_w_shots[i,] %>% pull('shot_loc_x')
          end_loc_y = pbp_w_shots[i,] %>% pull('shot_loc_y')
          end_shot_distance = 0
          end_shot_zone = ''
          end_shot_type = ''
          end_shot_result = 0
          end_shot_was_tpa = -1
          end_player = ''
          sequences = sequences %>% 
            add_row(sequence_id=nrow(sequences),
                    offense_team_id = as.double(curr_team),
                    init_action = curr_init_action,
                    end_action = end_action,
                    success=as.numeric(end_success),
                    init_player_id = as.character(curr_init_player_id),
                    end_player_id = as.character(end_player),
                    init_play_loc_x = as.numeric(curr_init_play_loc_x),
                    init_play_loc_y = as.numeric(curr_init_play_loc_y),
                    end_play_loc_x= as.numeric(end_loc_x),
                    end_play_loc_y = as.numeric(end_loc_y),
                    shot_distance = as.numeric(end_shot_distance),
                    shot_zone = as.character(end_shot_zone),
                    shot_result = as.numeric(end_shot_result),
                    shot_type = as.character(end_shot_type),
                    shot_was_tpa = as.numeric(end_shot_was_tpa))
          curr_init_action = ''
        }
        else if(tolower(pbp_w_shots[i, 'description']) %>% str_detect('turnover')){
          end_action = 'turnover'
          end_success = 0
          end_loc_x = pbp_w_shots[i,] %>% pull('poss_loc_x')
          end_loc_x = pbp_w_shots[i,] %>% pull('poss_loc_y')
          end_shot_distance = -1
          end_shot_zone = -1
          end_shot_type = ''
          end_shot_result = 0
          end_shot_was_tpa = -1
          end_player = pbp_w_shots[i,] %>% pull('player1_id')
          sequences = sequences %>% 
            add_row(sequence_id=nrow(sequences),
                    offense_team_id = curr_team,
                    init_action = curr_init_action,
                    end_action = end_action,
                    success=as.numeric(end_success),
                    init_player_id = as.character(curr_init_player_id),
                    end_player_id = as.character(end_player),
                    init_play_loc_x = as.numeric(curr_init_play_loc_x),
                    init_play_loc_y = as.numeric(curr_init_play_loc_y),
                    end_play_loc_x= as.numeric(end_loc_x),
                    end_play_loc_y = as.numeric(end_loc_y),
                    shot_distance = as.numeric(end_shot_distance),
                    shot_zone = as.character(end_shot_zone),
                    shot_result = as.numeric(end_shot_result),
                    shot_type = as.character(end_shot_type),
                    shot_was_tpa = as.numeric(end_shot_was_tpa))
          curr_init_action = ''
        }
      }}, error = function(e){
        curr_init_action = ''
        curr_team = ''
        curr_init_player_id = ''
        curr_init_play_loc_x = 0
        curr_init_play_loc_y = 0
      })
  }  
  
  return(sequences)
}

get_sequence_data_by_chunk = function(chunk){
  print('Worker started')
  sequences = get_pbp_game_sequence_data(game_i=chunk[1]) %>% mutate(game_id = chunk[1])
  for(i in 2:length(chunk)){
    tryCatch({
      s = get_pbp_game_sequence_data(game_i=chunk[i])
      sequences = sequences %>% rbind(s %>% mutate(game_id = chunk[i]))
      print(str_c("Done with game id ", as.character(chunk[i]), " ", as.character(i), " of ", as.character(length(chunk)), " (", as.character(i/length(chunk)), ")"))
      print(sequences %>% tail(5))
    }, error = function(e){})
  }
  return(sequences)
}

get_sequence_data_for_season = function(season){
  games = hoopR::nba_leaguegamelog(season=year_to_season(season))$LeagueGameLog 
  games = games %>% pull(GAME_ID) %>% unique()
  num_cores = detectCores()-1
  chunks = split(games, ceiling(seq_along(games)/num_cores))
  print('Beginning threading to process data')
  results = mclapply(chunks,get_sequence_data_by_chunk, mc.cores=num_cores)
  results = do.call(rbind, results)
  return(results)
}

d = get_sequence_data_for_season(SEASON_NAME)
d %>% write.csv('./SEQUENCES.csv')
```

```{r 2024 sequences, echo=FALSE}
sequences = read_csv('../data/ALL_SEQUENCES_2024.csv')
sequences %>% head(10)
```

### Understanding the dataset
```{r filtering}
sequences_w_players = sequences %>% filter(!is.na(init_player_id) & !is.na(end_player_id)) 
sequences_w_players %>% head(10)
```

```{r}
summary = sequences_w_players %>% group_by(init_action, end_action) %>% summarise(
  occurences = n(),  
  successes = sum(success)
) %>% arrange(desc(successes))
summary
```

```{r initiating actions}
init_actions = sequences_w_players %>% group_by(init_action) %>% summarise(
  occurences = n(),  
  successes = sum(success)
) %>% arrange(desc(successes))
init_actions
```

### Creating a simplistic Bayesian expectation model based on group means
```{r}
## Pr(success | init_action)
pr_success_init_action = sequences_w_players %>% group_by(init_action) %>% summarise(
  proba = sum(success)/n(),
) %>% arrange(desc(proba))
pr_success_init_action
```

```{r}
ggplot(pr_success_init_action, aes(init_action, proba)) + 
  geom_col() +
  labs(x="Initiating action", y="Probabiltiy of successful sequence", title="Probability Sequence Ends in Success Given Initiating Action")
```

```{r estimated initating values}
expectation_init_action_three = sequences_w_players %>% filter(end_action == 'shot' & shot_was_tpa >= 1) %>% group_by(init_action, shot_was_tpa) %>% summarise(
  proba = sum(success)/n(),
  occurences = n(),
  exp_val = (sum(success)/n())*3
) %>% arrange(desc(exp_val))

expectation_init_action_two = sequences_w_players %>% filter(end_action == 'shot' & shot_was_tpa == 0) %>% group_by(init_action, shot_was_tpa) %>% summarise(
  proba = sum(success)/n(),
  occurences = n(),
  exp_val = (sum(success)/n())*2
) %>% arrange(desc(exp_val))

summary_init_action = expectation_init_action_three %>% rbind(expectation_init_action_two) %>% arrange(exp_val)
summary_init_action
```

```{r}
summary_init_action %>% group_by(init_action) %>% summarise(total_xP = sum(exp_val)) %>% arrange(desc(total_xP))
```

### Using glm to estimate probaility of successful possession
```{r glm}
### Using init_action, coordinates to estimate probability of success
encoded_actions = sequences_w_players %>% mutate(
  init_action = case_when(
    init_action == 'inbound' ~ 0,
    init_action == 'rebound' ~ 1,
    init_action == 'steal' ~ 2,
    init_action == 'violation' ~ 3,
    .default = 0
  ),
  end_action = case_when(
    end_action == 'turnover' ~ 0,
    end_action == 'shot' ~ 1,
    .default = 0
  ),
  init_play_loc_x = init_play_loc_x/max(sequences_w_players$init_play_loc_x),
  init_play_loc_y = init_play_loc_y/max(sequences_w_players$init_play_loc_y)
)
prob_success = glm(success ~ init_action + init_play_loc_x + init_play_loc_y, family=binomial,data = encoded_actions)
summary(prob_success)
```

```{r testing glm}
data_prob_success = encoded_actions %>% select(c(init_player_id, init_action, init_play_loc_x, init_play_loc_y, success, shot_was_tpa))
test_set_prob_success = data_prob_success %>% sample_frac(size=0.25, replace=FALSE)
pred_success = predict(prob_success, newdata = test_set_prob_success, type='response')
test_set_prob_success = test_set_prob_success %>% mutate(pred_success = pred_success)
test_set_prob_success
```

```{r}
num_shot_types = encoded_actions %>% filter(shot_was_tpa >= 0 & success == 1) %>% group_by(success, shot_was_tpa) %>% summarise(
  occurences = n()
)
num_shot_types
```

```{r}
## E[Pr(shot_type | success)]
num_shot_types %>% filter(shot_was_tpa >= 0) %>% mutate(
  pot_val = c(2, 3),
  exp_val = pot_val*(num_shot_types[num_shot_types$shot_was_tpa == shot_was_tpa,] %>% pull('occurences'))/sum(num_shot_types$occurences)
)
```

```{r}
ev_glm = test_set_prob_success %>% mutate(init_eV = (1.36*pred_success+0.967*pred_success)) ## calculate expected value
ev_glm %>% head(10)
```

```{r best initiators in 2024}
top_10_initiators = ev_glm %>% group_by(init_player_id) %>% summarise(total_init_eV = sum(init_eV), sequences = n(), eV_per_sequence = sum(init_eV)/n()) %>% arrange(desc(total_init_eV))
top_10_initiators %>% 
  left_join(hoopR::nba_commonallplayers(season=year_to_season(2024))$CommonAllPlayers %>%
              rename(init_player_id = PERSON_ID) %>%
              mutate(init_player_id = as.numeric(init_player_id)) %>% select(c('init_player_id', 'DISPLAY_FIRST_LAST'))) %>%
  head(10) %>% rename(player_name = DISPLAY_FIRST_LAST) %>% select(c('player_name', everything()))
```

## Expected value based on end of possession
```{r}
data_eV_end = encoded_actions %>% select(c('success','end_action','end_player_id', 'end_play_loc_x', 'end_play_loc_y', 'shot_distance', 'shot_zone', 'shot_distance', 'shot_result', 'shot_type', 'shot_was_tpa'))
data_eV_end_shots = data_eV_end %>% filter(!is.na(shot_type) & shot_was_tpa >= 0)
```

```{r mapping locations to shot zones, echo=FALSE}
## Mapping possessions that did not end in shot zone to their given shot zone based on X,Y
shot_map = data_eV_end_shots %>% 
  group_by(shot_zone) %>% 
  summarise(min_x = min(end_play_loc_x), max_x = max(end_play_loc_x), min_y = min(end_play_loc_y), max_y = max(end_play_loc_y))

for(i in 1:nrow(data_eV_end)){
  if(data_eV_end[i,'shot_zone'] == '-1'){
    print(i)
    curr_x = data_eV_end[i,] %>% pull(end_play_loc_x)
    curr_y = data_eV_end[i,] %>% pull(end_play_loc_y)
    shot_zone = 'Center(C)'
    for(j in 1:nrow(shot_map)){
      min_x = shot_map[j,] %>% pull('min_x')
      max_x = shot_map[j,] %>% pull('max_x')
      min_y = shot_map[j,] %>% pull('min_y')
      max_y = shot_map[j,] %>% pull('max_y')
      if(curr_x >= min_x & curr_x <= max_x & curr_y >= min_y & curr_y <= max_y){
        shot_zone = shot_map[j,] %>% pull('shot_zone')
        break
      }
    }
    data_eV_end[i, 'shot_zone'] = shot_zone
  }
}
```

```{r estimating distance}
## Estimating missing distances based on coordinates
shot_distances = data_eV_end %>% filter(shot_distance >= -1)
distance_model = lm(shot_distance ~ 1+end_play_loc_y, data=shot_distances)
dist_predictions = predict(distance_model, newdata=data_eV_end)
```

```{r}
data_eV_end = data_eV_end %>% mutate(dist_pred = dist_predictions)
data_eV_end = data_eV_end %>% mutate(
  shot_distance = case_when(
    shot_distance < 0 ~ dist_pred,
    .default = shot_distance
  )
)
data_eV_end = data_eV_end %>% select(-c('dist_pred'))
data_eV_end = data_eV_end %>% mutate(shot_distance = round(shot_distance))
```

```{r}
## Encode shot_zones to numeric
encoded_data_eV_end = data_eV_end %>% mutate(
  shot_zone = case_when(
    shot_zone == 'Back Court(BC)' ~ 0,
    shot_zone == 'Center(C)' ~ 1,
    shot_zone == 'Left Side Center(LC)' ~ 2,
    shot_zone == 'Left Side(L)' ~ 3,
    shot_zone == 'Right Side Center(RC)' ~ 4,
    shot_zone == 'Right Side (R)' ~ 5,
    .default = 0
  )
)

max_x = max(encoded_data_eV_end$end_play_loc_x)
max_y = max(encoded_data_eV_end$end_play_loc_y)
max_distance = max(encoded_data_eV_end$shot_distance)
encoded_data_eV_end = encoded_data_eV_end %>% mutate(
  end_play_loc_x = end_play_loc_x/max_x,
  end_play_loc_y = end_play_loc_y/max_y,
  shot_distance = shot_distance/max_distance,
  shot_zone = shot_zone/5
) 
```

### Random Forest model for estimating the probability of a shot being made
```{r RF proba of shot made}
xg_model = ranger::ranger(success ~ shot_distance + end_play_loc_x + end_play_loc_y + shot_zone, data=encoded_data_eV_end, probability=TRUE, write.forest = TRUE)
shot_made_predictions = predict(xg_model, data = encoded_data_eV_end)

predictions = shot_made_predictions$predictions[,2]
data_eV_end = data_eV_end %>% mutate(
  pred_shot_made_prob = predictions
)

test_shot_prob_set = data_eV_end
three_pt_distance_by_zone = data_eV_end %>% filter(shot_was_tpa >= 1) %>% group_by(shot_zone) %>% summarise(distance = min(shot_distance))
test_shot_prob_set = test_shot_prob_set %>% mutate(in_three_range = numeric(length=nrow(test_shot_prob_set)))
for(i in 1:nrow(data_eV_end)){
  if(test_shot_prob_set[i,'shot_result'] <= -1){
    print(i)
    test_shot_prob_set[i, 'in_three_range'] = test_shot_prob_set[i, 'shot_distance'] %>% pull() >= three_pt_distance_by_zone[three_pt_distance_by_zone$shot_zone == test_shot_prob_set[i,'shot_zone'] %>% pull(), 'distance'] %>% pull()
    
  }
}
```

```{r}
test_shot_prob_set = test_shot_prob_set %>% mutate(
  xPts = case_when(
    shot_was_tpa > 0 | in_three_range > 0 ~ 3*pred_shot_made_prob,
    .default = 3*pred_shot_made_prob
  )
) 

test_shot_prob_set %>% group_by(end_player_id) %>% summarise(
  tot_xPts = sum(xPts)
) %>% arrange(desc(tot_xPts))
```

### Results of applying model across all sequences, by player
```{r}
xPts_w_players = test_shot_prob_set %>% group_by(end_player_id) %>% summarise(
  tot_xPts = sum(xPts), sequences = n(), xPts_per_sequence = sum(xPts)/n()
) %>% arrange(desc(tot_xPts)) %>% left_join(hoopR::nba_commonallplayers(season=year_to_season(2024))$CommonAllPlayers %>%
                                                rename(end_player_id = PERSON_ID) %>%
                                                mutate(end_player_id = as.numeric(end_player_id)) %>% select(c('end_player_id', 'DISPLAY_FIRST_LAST')))
xPts_w_players %>% head(10)
```

### Total xT (xPts of end + eV at beginning) by player
```{r}
init_players = top_10_initiators %>% rename(player_id = init_player_id, init_sequences = sequences) %>% select(c('player_id', 'total_init_eV', 'eV_per_sequence', 'init_sequences'))
end_players = xPts_w_players %>% rename(player_id = end_player_id, ended_sequences = sequences) %>% select(c('player_id', 'tot_xPts','xPts_per_sequence', 'DISPLAY_FIRST_LAST', 'ended_sequences'))
xT = init_players %>% left_join(end_players) %>% select(c('DISPLAY_FIRST_LAST', 'player_id', everything())) %>% rename(player_name = DISPLAY_FIRST_LAST)
top_20_xT_per_seq_min_1000_seq = xT %>% group_by(player_name) %>% filter(init_sequences + ended_sequences > 1000) %>% summarise(xT_per_sequence = xPts_per_sequence + eV_per_sequence) %>% arrange(desc(xT_per_sequence)) %>% head(20)

ggplot(top_20_xT_per_seq_min_1000_seq, aes(player_name, xT_per_sequence)) + 
  geom_col() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Player", y= "Total xT in 2024", title = "Top 20 Total xT Contributors in 2024")
  
top_20_xT_per_seq_min_1000_seq
```

```{r}
top_20_xT = xT %>% group_by(player_name) %>% summarise(tot_xT = tot_xPts + total_init_eV) %>% arrange(desc(tot_xT)) %>% head(20)
ggplot(top_20_xT, aes(player_name, tot_xT)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Player", y= "xT per Sequence in 2024", title = "Top 20 xT per Sequence Contributors in 2024", subtitle="Min. 1000 sequences")
top_20_xT
```


